Weighted Model Counting with Conditional Weights

### 2020-07-31 paper/paper.tex
Improvements to notation

\begin{itemize}
\item The Main Narrative
  \begin{enumerate}
  \item When weights are defined on literals, the measure on the free BA is
    fully independent.
  \item This means that the BA itself must be larger (i.e., have additional
    `meaningless' literals) to turn any probability distribution into an
    independent one.
  \item We show how we can define conditional weights on literals, allowing us
    to encode any probability distribution into a Boolean algebra that's not
    necessarily independent and thus can be smaller.
  \item We demonstrate a specific example of this by presenting a new way to
    encode Bayesian networks into instances of WMC and adapting a WMC algorithm
    (ADDMC) to run on the new format.
  \item We show that this results in significantly faster inference.
  \item We show that our encoding results in asymptotically fewer literals and
    fewer ADDs.
  \end{enumerate}
\item Introduce
  \begin{itemize}
  \item WMC
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} and ADDs
    \cite{DBLP:journals/fmsd/BaharFGHMPS97}
  \item A Bayesian network is a directed acyclic graph with random variables as
    vertices that defines a probability distribution over them. The full
    probability distribution can be expressed as the product of the conditional
    probability distributions of all random variables. For discrete Bayesian
    networks (and we only consider discrete networks in this paper), each such
    distribution can be represented by a table known as a conditional
    probability table (CPT).
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow.
\item Notable previous/related work
  \begin{itemize}
  \item Hailperin's approach to probability logic
    \cite{DBLP:journals/ndjfl/Hailperin84}
  \item Nilsson's (somewhat successful) probabilistic logic
    \cite{DBLP:journals/ai/Nilsson86,DBLP:journals/ai/Nilsson93}
  \item Measures on Boolean algebras
    \begin{itemize}
    \item On possibility and probability measures in finite Boolean algebras
      \cite{DBLP:journals/soco/CastineiraCT02}
    \item Representation of conditional probability measures
      \cite{krauss1968representation}
    \end{itemize}
  \end{itemize}
\item Our work is, in spirit, similar to... using ROBDDs for Inference in
  Bayesian Networks with Troubleshooting as an Example
  \cite{DBLP:conf/uai/NielsenWJK00}: this is an interesting approach where
  deterministic parts of a BN are expressed as Boolean functions, i.e.,
  extracting the logical from the probabilistic. Maybe I can make a case that my
  encoding (the textual version) kind of does the same, even though the
  algorithm doesn't.
\end{itemize}

@@

\paragraph{Using WMC to perform inference on Bayesian networks.} Hitherto, four
techniques have been proposed for encoding Bayesian networks into instances of
WMC. We will identify them based on the initials of authors as well as
publications years: \texttt{d02} \cite{DBLP:conf/kr/Darwiche02}, \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05}, \texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05},
and \texttt{cd06} \cite{DBLP:conf/sat/ChaviraD06}. Below we friendly summarise
the observed performance differences among them. Sang et al.
\cite{DBLP:conf/aaai/SangBK05} claim that \texttt{sbk05} is a smaller encoding
than \texttt{d02} with respect to both the number of clauses and the number of
variables but provide no experimental comparison. Chavira and Darwiche
\cite{DBLP:conf/ijcai/ChaviraD05} compare \texttt{cd05} with \texttt{d02} by
measuring the time it takes to compile either encoding into an arithmetic
circuit (but do not measure inference time). The more recent encoding
\texttt{cd05} always compiles faster and results in a smaller arithmetic circuit
(as measured by the number of edges). In their subsequent paper, the same
authors perform two sets of experiments (that are relevant to this summary)
\cite{DBLP:conf/sat/ChaviraD06}. First, they compile \texttt{cd05} and
\texttt{cd06} encodings into d-DNNF (i.e., deterministic decomposable negation
normal form \cite{DBLP:journals/jancl/Darwiche01}), measuring both compilation
time and numbers of edges in the d-DNNF diagram. The results are mostly in
favour of \texttt{cd06}. Second, they compare the inference time of
\texttt{sbk05} run with Cachet \cite{DBLP:conf/sat/SangBBKP04} with the compile
times of \texttt{cd05} and \texttt{cd06}, but only on five (types of) instances.
In these experiments, \texttt{cd06} is always faster than \texttt{cd05}, while
the comparison with \texttt{sbk05} is mixed. Sometimes \texttt{cd06} is orders
of magnitude faster than \texttt{sbk05}, sometimes slightly slower. The
performance difference between \texttt{sbk05} and \texttt{cd05} is even harder
to judge: \texttt{sbk05} is better on three out of five instances and worse on
the remaining two. Based on this description, one would expect \texttt{cd06} to
be faster than both \texttt{cd05} and \texttt{sbk05}, both of which should be
faster than \texttt{d02}. The experiments in \cref{sec:experiments}, however,
strongly disagree with this prediction, showing that the quality of an encoding
depends strongly on the underlying search algorithm or compilation technique.

\paragraph{ADDs and their use in probabilistic inference.} ADDs provide an
efficient way to manipulate functions from a Boolean algebra to any algebraic
structure (most commonly $\mathbb{R}$) \cite{DBLP:journals/fmsd/BaharFGHMPS97}.
They have been used to represent value functions in Markov decision processes
\cite{DBLP:conf/uai/HoeySHB99} and, for Bayesian network inference, to represent
each CPT as an ADD \cite{DBLP:conf/icml/ZhaoMP15} as well as by combining ADDs
and arithmetic circuits into a  single representation
\cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous in this
situation because of their ability to fully exploit context-specific
independence, i.e., observable structure within a CPT that is not inherited from
the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

@@

\begin{definition}[Operations on functions]
  Let $A\colon 2^X \to \mathbb{R}_{\ge 0}$ and $B\colon 2^Y \to \mathbb{R}_{\ge
    0}$ be functions, $\alpha \in \mathbb{R}_{\ge 0}$, and $x \in X$. We define
  the following operations:
  \begin{description}
  \item[Addition:] $A+B\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$ is
    such that $(A+B)(\tau) = A(\tau \cap X) + B(\tau \cap Y)$ for all $\tau \in
    2^{X \cup Y}$.
  \item[Multiplication:] $A \cdot B\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$
    is such that $(A \cdot B)(\tau) = A(\tau \cap X) \cdot B(\tau
    \cap Y)$ for all $\tau \in 2^{X \cup Y}$.
  \item[Scalar multiplication:] $\alpha A\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $(\alpha A)(\tau) = \alpha \cdot A(\tau)$ for all $\tau
    \in 2^X$.
  \item[Complement:] $\overline{A}\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $\overline{A}(\tau) = 1 - A(\tau)$ for all $\tau \in 2^X$.
  \item[Projection:] $\exists_xA\colon 2^{X \setminus \{ x \}} \to
    \mathbb{R}_{\ge 0}$ is such that $(\exists_xA)(\tau) = A(\tau) +
    A(\tau \cup \{ x \})$ for all $\tau \in 2^{X \setminus \{x \}}$.
  \end{description}
\end{definition}

\begin{observation}
  Let $U$ be a set, and $\mathcal{V} = \{ A\colon 2^X \to \mathbb{R}_{\ge 0}
  \mid X \subseteq U \}$. Then $\mathcal{V}$ is a semi-vector space with three
  additional operations: inverse, (non-scalar) multiplication, and projection.
  Specifically, note that both addition and multiplication are both associative
  and commutative.
\end{observation}

@@

Henceforth, for any function $A\colon 2^X \to \mathbb{R}_{\ge 0}$ and any set
$\tau$, we will write $A(\tau)$ to mean $A(\tau \cap X)$.

@@

Let $U$ be a set. A \emph{measure} is a function $M\colon 2^{2^U} \to
\mathbb{R}_{\ge 0}$ such that $M(\bot) = 0$, and $M(a \lor b) = M(a) + M(b)$ for
all $a, b \in 2^{2^U}$ whenever $a \land b = \bot$. A \emph{weight function} is
a function $W\colon 2^U \to \mathbb{R}_{\ge 0}$. A weight function is
\emph{factored} if $W = \prod_{x \in U} W_x$ for some functions $W_x\colon
2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$. We say that a weight function
$W\colon 2^U \to \mathbb{R}_{\ge 0}$ \emph{induces} a measure $M_W\colon 2^{2^U}
\to \mathbb{R}_{\ge 0}$ if
\begin{equation} \label{eq:induced_measure}
  M_W(x) = \sum_{\{u\} \le x} W(u).
\end{equation}
Finally, a measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is \emph{factorable}
if there exists a factored weight function $W\colon 2^U \to \mathbb{R}_{\ge 0}$
that induces $M$.

\begin{lemma} \label{prop:measure}
  The function $M_W$, as defined by \cref{eq:induced_measure}, is a measure.
\end{lemma}
\begin{proof}
  Note that $M_W(\bot) = 0$ since there are no atoms below $\bot$. Let $a, b \in
  2^{2^{U}}$ be such that $a \land b = \bot$. By elementary properties of
  Boolean algebras, all atoms below $a \lor b$ are either below $a$ or below
  $b$. Moreover, none of them can be below both $a$ and $b$ because then they
  would have to be below $a \land b = \bot$. Thus
  \[
    M_W(a \lor b) = \sum_{\{u\} \le a \lor b} W(u) = \sum_{\{u\} \le a} W(u) +
    \sum_{\{u\} \le b} = M_W(a) + M_W(b)
  \]
  as required.
\end{proof}

In this formulation, the process of calculating the value of $M_W(x)$ for some
$x \in 2^{2^U}$ with a given definition of $W$ is known as WMC.

\paragraph{Relation to the classical (logic-based) view of WMC.} Let
$\mathcal{L}$ be a propositional logic with two atoms $a$ and $b$ as in
\cref{sec:prelims} and $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge
  0}$ a \emph{weight function} defined as $w(a) = 0.3$, $w(\neg a) = 0.7$, $w(b)
= 0.2$, $w(\neg b) = 0.8$. Furthermore, let $\Delta$ be a theory in
$\mathcal{L}$ with a sole axiom $a$. Then $\Delta$ has two models: $\{ a, b \}$
and $\{ a, \neg b \}$ and its WMC \cite{DBLP:journals/ai/ChaviraD08} is
\begin{equation} \label{eq:wmc_example}
  \mathrm{WMC}(\Delta) = \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) = w(a)w(b) + w(a)w(\neg b) = 0.3.
\end{equation}
Alternatively, we can define $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
$W_a(\{ a \}) = 0.3$, $W_a(\emptyset) = 0.7$ and $W_b\colon 2^{\{q\}} \to
\mathbb{R}_{\ge 0}$ as $W_b(\{ b \}) = 0.2$, $W_b(\emptyset) = 0.8$. Let $M$ be
the measure on $2^{2^U}$ induced by $W = W_a \cdot W_b$. Then, equivalently to
\cref{eq:wmc_example}, we can write
\[
  M(a) = W(\{ a, b \}) + W(\{ a \}) = W_a(\{a\})W_b(\{b\}) +
  W_a(\{a\})W_b(\emptyset) = 0.3.
\]

@@

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $M\colon 2^{2^U} \to \mathbb{R}_{\ge
    0}$ a measure defined as:\footnote{The value of $M$ on any other element of
    the Boolean algebra can be deduced using the definition.}
  \begin{align*}
    M(a \land b) &= 0.72, \\
    M(a \land \neg b) &= 0.18, \\
    M(\neg a \land b) &= 0.07, \\
    M(\neg a \land \neg b) &= 0.03.
  \end{align*}
  If $M$ could be represented using traditional (factored) WMC, we would have to
  find two weight functions $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and
  $W_b\colon 2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $W = W_a \cdot W_b$
  induces $M$, i.e., $W_a$ and $W_b$ would have to satisfy this system of
  equations:
  \begin{align*}
    W_a(\{a\}) \cdot W_b(\{b\}) &= 0.72 \\
    W_a(\{a\}) \cdot W_b(\emptyset) &= 0.18 \\
    W_a(\emptyset) \cdot W_b(\{b\}) &= 0.07 \\
    W_a(\emptyset) \cdot W_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solutions.

  Alternatively, we can let $b$ depend on $a$ and consider weight functions
  $W_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $W_b\colon 2^{\{a, b\}} \to
  \mathbb{R}_{\ge 0}$ defined as $W_a(\{a\}) = 0.9$, $W_a(\emptyset) = 0.1$, and
  $W_b(\{a, b\}) = 0.8$, $W_b(\{a\}) = 0.2$, $W_b(\{b\}) = 0.7$, $W_b(\emptyset)
  = 0.3$. One can easily check that with these definitions $W$ indeed induces
  $M$.
\end{example}

Note that in this case we chose to interpret $W_b$ as $\Pr(B \mid A)$
while---with a different definition of $W_b$ that represents the joint
probability distribution $\Pr(A, B)$---$W_b$ by itself could induce $M$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage \cite{DBLP:conf/aaai/DudekPV20}.

\begin{lemma} \label{lemma:before_theorem}
  For any measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ and elements $a, b
  \in 2^{2^U}$,
  \begin{equation} \label{eq:to_prove}
    M(a \land b) = M(a)M(b)
  \end{equation}
  if and only if
  \begin{equation} \label{eq:to_prove2}
    M(a \land b) \cdot M(\neg a \land \neg b) = M(a \land \neg b)
    \cdot M(\neg a \land b).
  \end{equation}
\end{lemma}
\begin{proof}
  First, note that $a = (a \land b) \lor (a \land \neg b)$ and $(a \land b)
  \land (a \land \neg b) = 0$, so, by properties of a measure,
  \begin{equation} \label{eq:temp}
    M(a) = M(a \land b) + M(a \land \neg b).
  \end{equation}
  Applying \cref{eq:temp} and the equivalent expression for $M(b)$ allows us
  to rewrite \cref{eq:to_prove} as
  \[
    M(a \land b) = [M(a \land b) + M(a \land \neg b)][M(a \land b) + M(\neg a
    \land b)]
  \]
  which is equivalent to
  \begin{equation} \label{eq:temp6}
    M(a \land b)[1 - M(a \land b) - M(a \land \neg b) - M(\neg a \land b)] = M(a
    \land \neg b)M(\neg a \land b).
  \end{equation}
  Since $a \land b$, $a \land \neg b$, $\neg a \land b$, $\neg a \land \neg b$
  are pairwise disjoint and their supremum is $1$,
  \[
    M(a \land b) + M(a \land \neg b) + M(\neg a \land b) + M(\neg a \land \neg
    b) = 1,
  \]
  and this allows us to rewrite \cref{eq:temp6} into \cref{eq:to_prove2}. As all
  transformations are invertible, the two expressions are equivalent.
\end{proof}

\begin{theorem}
  A measure $M\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is factorable if and only
  if
  \begin{equation} \label{eq:wmccondition}
  M(u \land v) = M(u)M(v)
  \end{equation}
  for all distinct $u, v \in U \cup \{ \neg w \mid w \in U \}$ such that $u \ne
  \neg v$.
\end{theorem}
\begin{proof}
  ($\Leftarrow$) For each $x \in U$, let $W_x\colon 2^{\{x\}} \to
  \mathbb{R}_{\ge 0}$ be defined by $W_x(\{ x \}) = M(x)$ and $W_x(\emptyset) =
  M(\neg x)$. Let $M_W$ be the measure induced by $W = \prod_{x \in U} W_x$. We
  will show that $M = M_W$. First, note that $M_w(\bot) = 0 = M(\bot)$ by
  \cref{prop:measure} and the definition of a measure. Second, let $a =
  \bigwedge_{u \in U} a_u$ be an atom in $2^{2^U}$ such that $a_u \in \{ u, \neg
  u \}$ for all $u \in U$. Then
  \[
    M_W(a) = W(\{ a_u \mid u \in U \}) = \prod_{u \in U} W_u(\{a_u\}) = \prod_{u
      \in U} M(a_u) = M \left(\bigwedge_{u \in U} a_u \right) = M(a)
  \]

  Finally, note that if $M_W$ and
  $M$ agree on all atoms, then they must also agree on all other non-zero
  elements of the Boolean algebra.

>>>>

\begin{itemize}
\item The Main Narrative
  \begin{enumerate}
  \item When weights are defined on literals, the measure on the free BA is
    fully independent.
  \item This means that the BA itself must be larger (i.e., have additional
    `meaningless' literals) to turn any probability distribution into an
    independent one.
  \item We show how we can define conditional weights on literals, allowing us
    to encode any probability distribution into a Boolean algebra that's not
    necessarily independent and thus can be smaller.
  \item We demonstrate a specific example of this by presenting a new way to
    encode Bayesian networks into instances of WMC and adapting a WMC algorithm
    (ADDMC) to run on the new format.
  \item We show that this results in significantly faster inference.
  \item We show that our encoding results in asymptotically fewer literals and
    fewer ADDs.
  \end{enumerate}
\item Introduce
  \begin{itemize}
  \item WMC \cite{DBLP:conf/aaai/SangBK05}
  \item ADDMC \cite{DBLP:conf/aaai/DudekPV20} and ADDs
    \cite{DBLP:journals/fmsd/BaharFGHMPS97}
  \item A Bayesian network is a directed acyclic graph with random variables as
    vertices that defines a probability distribution over them. The full
    probability distribution can be expressed as the product of the conditional
    probability distributions of all random variables. For discrete Bayesian
    networks (and we only consider discrete networks in this paper), each such
    distribution can be represented by a table known as a conditional
    probability table (CPT).
  \end{itemize}
\item[F] What are the main claims, what are the main takeaways, intuitive [???]
  of theorems to follow.
\item Our work is, in spirit, similar to... using ROBDDs for Inference in
  Bayesian Networks with Troubleshooting as an Example
  \cite{DBLP:conf/uai/NielsenWJK00}: this is an interesting approach where
  deterministic parts of a BN are expressed as Boolean functions, i.e.,
  extracting the logical from the probabilistic. Maybe I can make a case that my
  encoding (the textual version) kind of does the same, even though the
  algorithm doesn't.
\end{itemize}

@@

\paragraph{Using WMC to perform inference on Bayesian networks.} Hitherto, four
techniques have been proposed for encoding Bayesian networks into instances of
WMC. We will identify them based on the initials of authors as well as
publications years: \texttt{d02} \cite{DBLP:conf/kr/Darwiche02}, \texttt{sbk05}
\cite{DBLP:conf/aaai/SangBK05}, \texttt{cd05} \cite{DBLP:conf/ijcai/ChaviraD05},
and \texttt{cd06} \cite{DBLP:conf/sat/ChaviraD06}. Below we summarise the
observed performance differences among them. Sang et al.
\cite{DBLP:conf/aaai/SangBK05} claim that \texttt{sbk05} is a smaller encoding
than \texttt{d02} with respect to both the number of clauses and the number of
variables but provide no experimental comparison. Chavira and Darwiche
\cite{DBLP:conf/ijcai/ChaviraD05} compare \texttt{cd05} with \texttt{d02} by
measuring the time it takes to compile either encoding into an arithmetic
circuit (but do not measure inference time). The more recent encoding
\texttt{cd05} always compiles faster and results in a smaller arithmetic circuit
(as measured by the number of edges). In their subsequent paper, the same
authors perform two sets of experiments (that are relevant to this summary)
\cite{DBLP:conf/sat/ChaviraD06}. First, they compile \texttt{cd05} and
\texttt{cd06} encodings into d-DNNF (i.e., deterministic decomposable negation
normal form \cite{DBLP:journals/jancl/Darwiche01}), measuring both compilation
time and numbers of edges in the d-DNNF diagram. The results are mostly in
favour of \texttt{cd06}. Second, they compare the inference time of
\texttt{sbk05} run with Cachet \cite{DBLP:conf/sat/SangBBKP04} with the compile
times of \texttt{cd05} and \texttt{cd06}, but only on five (types of) instances.
In these experiments, \texttt{cd06} is always faster than \texttt{cd05}, while
the comparison with \texttt{sbk05} is mixed. Sometimes \texttt{cd06} is orders
of magnitude faster than \texttt{sbk05}, sometimes slightly slower. The
performance difference between \texttt{sbk05} and \texttt{cd05} is even harder
to judge: \texttt{sbk05} is better on three out of five instances and worse on
the remaining two. Based on this description, one would expect \texttt{cd06} to
be faster than both \texttt{cd05} and \texttt{sbk05}, both of which should be
faster than \texttt{d02}. The experiments in \cref{sec:experiments}, however,
strongly disagree with this prediction, showing that the quality of an encoding
depends strongly on the underlying search algorithm or compilation technique.

\paragraph{ADDs and their use in probabilistic inference.} ADDs provide an
efficient way to manipulate functions from a Boolean algebra to any algebraic
structure (most commonly the real numbers)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. They have been used to represent value
functions in Markov decision processes \cite{DBLP:conf/uai/HoeySHB99} and, for
Bayesian network inference, to represent each CPT as an ADD
\cite{DBLP:conf/icml/ZhaoMP15} as well as by combining ADDs and arithmetic
circuits into a  single representation \cite{DBLP:conf/ijcai/ChaviraD07}. ADDs
are particularly advantageous in this situation because of their ability to
fully exploit context-specific independence, i.e., observable structure within a
CPT that is not inherited from the structure of the Bayesian network
\cite{DBLP:conf/uai/BoutilierFGK96}.

@@

\begin{definition}[Operations on functions]
  Let $\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ and $\beta\colon 2^Y \to
  \mathbb{R}_{\ge 0}$ be functions, $p \in \mathbb{R}_{\ge 0}$, and $x \in X$.
  We define the following operations:
  \begin{description}
  \item[Addition:] $\alpha + \beta\colon 2^{X \cup Y} \to \mathbb{R}_{\ge 0}$ is
    such that $(\alpha + \beta)(T) = \alpha(T \cap X) + \beta(T \cap Y)$ for all
    $T \in 2^{X \cup Y}$.
  \item[Multiplication:] $\alpha \cdot \beta\colon 2^{X \cup Y} \to
    \mathbb{R}_{\ge 0}$ is such that $(\alpha \cdot \beta)(T) = \alpha(T \cap X)
    \cdot \beta(T \cap Y)$ for all $T \in 2^{X \cup Y}$.
  \item[Scalar multiplication:] $p\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $(p\alpha)(T) = p \cdot \alpha(T)$ for all $T \in 2^X$.
  \item[Complement:] $\overline{\alpha}\colon 2^X \to \mathbb{R}_{\ge 0}$ is
    such that $\overline{\alpha}(T) = 1 - \alpha(T)$ for all $T \in 2^X$.
  \item[Projection:] $\exists_x\alpha\colon 2^{X \setminus \{ x \}} \to
    \mathbb{R}_{\ge 0}$ is such that $(\exists_x\alpha)(T) = \alpha(T) +
    \alpha(T \cup \{ x \})$ for all $T \in 2^{X \setminus \{x \}}$.
  \end{description}
\end{definition}

\begin{observation}
  Let $U$ be a set, and $\mathcal{V} = \{ \alpha\colon 2^X \to \mathbb{R}_{\ge 0}
  \mid X \subseteq U \}$. Then $\mathcal{V}$ is a semi-vector space with three
  additional operations: (non-scalar) multiplication, complement, and projection.
  Specifically, note that both addition and multiplication are both associative
  and commutative.
\end{observation}

@@

Henceforth, for any function $\alpha\colon 2^X \to \mathbb{R}_{\ge 0}$ and any
set $T$, we will write $\alpha(T)$ to mean $\alpha(T \cap X)$.

@@

Let $U$ be a set. A \emph{measure} is a function $\mu\colon 2^{2^U} \to
\mathbb{R}_{\ge 0}$ such that $\mu(\bot) = 0$, and $\mu(a \lor b) = \mu(a) +
\mu(b)$ for all $a, b \in 2^{2^U}$ whenever $a \land b = \bot$. A \emph{weight
  function} is a function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$. A weight
function is \emph{factored} if $\nu = \prod_{x \in U} \nu_x$ for some functions
$\nu_x\colon 2^{\{x\}} \to \mathbb{R}_{\ge 0}$, $x \in U$. We say that a weight
function $\nu\colon 2^U \to \mathbb{R}_{\ge 0}$ \emph{induces} a measure
$\mu_\nu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ if
\begin{equation} \label{eq:induced_measure}
  \mu_\nu(x) = \sum_{\{u\} \le x} \nu(u).
\end{equation}
Finally, a measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is
\emph{factorable} if there exists a factored weight function $\nu\colon 2^U \to
\mathbb{R}_{\ge 0}$ that induces $\mu$.

\begin{lemma} \label{prop:measure}
  The function $\mu_\nu$, as defined by \cref{eq:induced_measure}, is a measure.
\end{lemma}
\begin{proof}
  Note that $\mu_\nu(\bot) = 0$ since there are no atoms below $\bot$. Let $a, b
  \in 2^{2^{U}}$ be such that $a \land b = \bot$. By elementary properties of
  Boolean algebras, all atoms below $a \lor b$ are either below $a$ or below
  $b$. Moreover, none of them can be below both $a$ and $b$ because then they
  would have to be below $a \land b = \bot$. Thus
  \[
    \mu_\nu(a \lor b) = \sum_{\{u\} \le a \lor b} \nu(u) = \sum_{\{u\} \le a}
    \nu(u) + \sum_{\{u\} \le b} \nu(u) = \mu_\nu(a) + \mu_\nu(b)
  \]
  as required.
\end{proof}

In this formulation, the process of calculating the value of $\mu_\nu(x)$ for
some $x \in 2^{2^U}$ with a given definition of $\nu$ is known as WMC.

\paragraph{Relation to the classical (logic-based) view of WMC.} Let
$\mathcal{L}$ be a propositional logic with two atoms $a$ and $b$ as in
\cref{sec:prelims} and $w\colon \{ a, b, \neg a, \neg b \} \to \mathbb{R}_{\ge
  0}$ a \emph{weight function} defined as $w(a) = 0.3$, $w(\neg a) = 0.7$, $w(b)
= 0.2$, $w(\neg b) = 0.8$. Furthermore, let $\Delta$ be a theory in
$\mathcal{L}$ with a sole axiom $a$. Then $\Delta$ has two models: $\{ a, b \}$
and $\{ a, \neg b \}$ and its WMC \cite{DBLP:journals/ai/ChaviraD08} is
\begin{equation} \label{eq:wmc_example}
  \mathrm{WMC}(\Delta) = \sum_{\omega \models \Delta} \prod_{\omega \models l} w(l) = w(a)w(b) + w(a)w(\neg b) = 0.3.
\end{equation}
Alternatively, we can define $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ as
$\nu_a(\{ a \}) = 0.3$, $\nu_a(\emptyset) = 0.7$ and $\nu_b\colon 2^{\{b\}} \to
\mathbb{R}_{\ge 0}$ as $\nu_b(\{ b \}) = 0.2$, $\nu_b(\emptyset) = 0.8$. Let
$\mu$ be the measure on $2^{2^U}$ induced by $\nu = \nu_a \cdot \nu_b$. Then,
equivalently to \cref{eq:wmc_example}, we can write
\[
  \mu(a) = \nu(\{ a, b \}) + \nu(\{ a \}) = \nu_a(\{a\})\nu_b(\{b\}) +
  \nu_a(\{a\})\nu_b(\emptyset) = 0.3.
\]

@@

\begin{example}
  Let $U = \{a, b\}$ be a set of atoms and $\mu\colon 2^{2^U} \to
  \mathbb{R}_{\ge 0}$ a measure defined as:\footnote{The value of $\mu$ on any
    other element of the Boolean algebra can be deduced using the definition.}
  \begin{align*}
    \mu(a \land b) &= 0.72, \\
    \mu(a \land \neg b) &= 0.18, \\
    \mu(\neg a \land b) &= 0.07, \\
    \mu(\neg a \land \neg b) &= 0.03.
  \end{align*}
  If $\mu$ could be represented using traditional (factored) WMC, we would have
  to find two weight functions $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$
  and $\nu_b\colon 2^{\{b\}} \to \mathbb{R}_{\ge 0}$ such that $\nu = \nu_a
  \cdot \nu_b$ induces $\mu$, i.e., $\nu_a$ and $\nu_b$ would have to satisfy
  this system of equations:
  \begin{align*}
    \nu_a(\{a\}) \cdot \nu_b(\{b\}) &= 0.72 \\
    \nu_a(\{a\}) \cdot \nu_b(\emptyset) &= 0.18 \\
    \nu_a(\emptyset) \cdot \nu_b(\{b\}) &= 0.07 \\
    \nu_a(\emptyset) \cdot \nu_b(\emptyset) &= 0.03,
  \end{align*}
  which has no solutions.

  Alternatively, we can let $b$ depend on $a$ and consider weight functions
  $\nu_a\colon 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ and $\nu_b\colon 2^{\{a, b\}}
  \to \mathbb{R}_{\ge 0}$ defined as $\nu_a(\{a\}) = 0.9$, $\nu_a(\emptyset) =
  0.1$, and $\nu_b(\{a, b\}) = 0.8$, $\nu_b(\{a\}) = 0.2$, $\nu_b(\{b\}) = 0.7$,
  $\nu_b(\emptyset) = 0.3$. One can easily check that with these definitions
  $\nu$ indeed induces $\mu$.
\end{example}

Note that in this case we chose to interpret $\nu_b$ as $\Pr(b \mid a)$
while---with a different definition of $\nu_b$ that represents the joint
probability distribution $\Pr(a, b)$---$\nu_b$ by itself could induce $\mu$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage \cite{DBLP:conf/aaai/DudekPV20}.

\begin{lemma} \label{lemma:before_theorem}
  For any measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ and elements $a, b
  \in 2^{2^U}$,
  \begin{equation} \label{eq:to_prove}
    \mu(a \land b) = \mu(a)\mu(b)
  \end{equation}
  if and only if
  \begin{equation} \label{eq:to_prove2}
    \mu(a \land b) \cdot \mu(\neg a \land \neg b) = \mu(a \land \neg b)
    \cdot \mu(\neg a \land b).
  \end{equation}
\end{lemma}
\begin{proof}
  First, note that $a = (a \land b) \lor (a \land \neg b)$ and $(a \land b)
  \land (a \land \neg b) = 0$, so, by properties of a measure,
  \begin{equation} \label{eq:temp}
    \mu(a) = \mu(a \land b) + \mu(a \land \neg b).
  \end{equation}
  Applying \cref{eq:temp} and the equivalent expression for $\mu(b)$ allows us
  to rewrite \cref{eq:to_prove} as
  \[
    \mu(a \land b) = [\mu(a \land b) + \mu(a \land \neg b)][\mu(a \land b) +
    \mu(\neg a \land b)]
  \]
  which is equivalent to
  \begin{equation} \label{eq:temp6}
    \mu(a \land b)[1 - \mu(a \land b) - \mu(a \land \neg b) - \mu(\neg a \land b)] = \mu(a \land \neg b)\mu(\neg a \land b).
  \end{equation}
  Since $a \land b$, $a \land \neg b$, $\neg a \land b$, $\neg a \land \neg b$
  are pairwise disjoint and their supremum is $\top$,
  \[
    \mu(a \land b) + \mu(a \land \neg b) + \mu(\neg a \land b) + \mu(\neg a
    \land \neg b) = 1,
  \]
  and this allows us to rewrite \cref{eq:temp6} into \cref{eq:to_prove2}. As all
  transformations are invertible, the two expressions are equivalent.
\end{proof}

\todo[inline]{The proof is only valid for probability measures!}

\begin{theorem}
  A measure $\mu\colon 2^{2^U} \to \mathbb{R}_{\ge 0}$ is factorable if and only
  if
  \begin{equation} \label{eq:wmccondition}
  \mu(u \land v) = \mu(u)\mu(v)
  \end{equation}
  for all distinct $u, v \in U \cup \{ \neg w \mid w \in U \}$ such that $u \ne
  \neg v$.
\end{theorem}
\begin{proof}
  ($\Leftarrow$) For each $x \in U$, let $\nu_x\colon 2^{\{x\}} \to
  \mathbb{R}_{\ge 0}$ be defined by $\nu_x(\{ x \}) = \mu(x)$ and
  $\nu_x(\emptyset) = \mu(\neg x)$. Let $\mu_\nu$ be the measure induced by $\nu
  = \prod_{x \in U} \nu_x$. We will show that $\mu = \mu_\nu$. First, note that
  $\mu_\nu(\bot) = 0 = \mu(\bot)$ by \cref{prop:measure} and the definition of a
  measure. Second, let $a = \bigwedge_{u \in U} a_u$ be an atom in $2^{2^U}$
  such that $a_u \in \{ u, \neg u \}$ for all $u \in U$. Then
  \[
    \mu_\nu(a) = \nu(\{ a_u \mid u \in U \}) = \prod_{u \in U} \nu_u(\{a_u\}) =
    \prod_{u \in U} \mu(a_u) = \mu\left(\bigwedge_{u \in U} a_u \right) = \mu(a)
  \]

  Finally, note that if $\mu_\nu$ and $\mu$ agree on all atoms, then they must
  also agree on all other non-zero elements of the Boolean algebra.
